---
title: "Live Session Unit 10 Assignment"
author: "Gino Varghese"
date: "November 11, 2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction              
Download click through rate data set from the following site   

* http://stat.columbia.edu/~rachel/datasets/nyt7.csv      
and perform necessary analysis to understand days worth of ads shown and clicks recorded on the New York Times homepage in May 2012, for this excercise i will be limiting my analysis to nyt7.csv file.         

<br>

#### The following was done before starting the analysis:                                

* setting of working directory                  
* installation and loading of necessary packages              
* R version     

```{r libsetup, include=TRUE, message=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/git/MSDS6306/LiveAssignment_Unit10/Analysis/Data")
require(knitr)
require(dplyr)
require(plyr)
require(ggplot2)
require(downloader)
sessionInfo()
```               
<br>

##### **The analysis was started by downloading the following data sets:**                                

* **download data set nyt7 dataset**         
```{r nyt7download, cache=TRUE,echo=TRUE}
source("downloadNYT7.R", echo = TRUE)
list.files()
```            
<br>

##### **Once the data is downloaded to the *Data* directory, its time to load the csv for analysis.**                

```{r loadnyt7, echo=TRUE}
filelocation <- "~/git/MSDS6306/LiveAssignment_Unit10/Analysis/Data/nyt7.csv"
nyt7data <- read.csv(filelocation)
```         
<br>

##### **Review the data set**         

```{r headnyt7, echo=TRUE}
head(nyt7data)
str(nyt7data)
summary(nyt7data)
```          
<br>

##### **Create a new variable *ageGroup* that categorizes age into following groups: < 18, 18–24, 25–34, 35–44, 45–54, 55–64 and 65+.**            

```{r ageGroupnyt7, echo=TRUE}
nyt7data$ageGroup <- cut(nyt7data$Age, c(-Inf, 18, 24, 34, 44, 54, 64, Inf))
levels(nyt7data$ageGroup) <- c("<18","18-24","25-34","35-44","45-54","55-64","65+")
levels(nyt7data$ageGroup)
head(nyt7data)
```          
<br>

##### **Use sub set of data called *ImpSub* where Impressions > 0  in your data set.**

```{r impsubnyt7, echo=TRUE}
ImpSub <- subset(nyt7data,Impressions > 0)
summary(ImpSub$Impressions)
```         
<br>

##### **Create a new variable called click-through-rate (CTR = click/impression).**           
```{r ctrimpsubnyt7, echo=TRUE}
ImpSub$CTR <- ImpSub$Clicks/ImpSub$Impressions
head(ImpSub)
```         
<br>

##### **Plot distributions of number impressions and click-through-rate (CTR = click/impression) for the age groups.**
<br>

```{r impsubplot1, echo=TRUE, cache=TRUE}
ggplot(subset(ImpSub,CTR>0), aes(x=CTR, fill = ageGroup)) +
  geom_histogram(binwidth = .05) +
  ggtitle("(CTR) Click Through Rate by Age Group") + xlab("CTR (Clicks/Impressions)") + ylab("Count")
``` 
<br>

```{r impsubplot2, echo=TRUE, cache=TRUE}
ggplot(subset(ImpSub,CTR>0), aes(x=log(CTR), fill = ageGroup)) +
  geom_histogram(binwidth = .05) +
  ggtitle("log(CTR) Click Through Rate by Age Group") + xlab("log CTR (Clicks/Impressions)") + ylab("Count")
```  
<br>

```{r impsubplot3, echo=TRUE, cache=TRUE}
ggplot(subset(ImpSub,CTR>0), aes(log(CTR), colour = ageGroup)) +
    geom_density(alpha=0.1,position = "stack") +
  ggtitle("log(CTR) Click Through Rate by Age Group") + xlab("CTR (Clicks/Impressions)") + ylab("Density")
```  
<br>   

##### **Define a new variable to segment users based on click -through-rate (CTR) behavior. CTR< 0.2, 0.2<=CTR <0.4, 0.4<= CTR<0.6, 0.6<=CTR<0.8, CTR>0.8**
```{r CTRbehavior, echo=TRUE, cache=TRUE}
ImpSub$CTR.behavior[ImpSub$CTR < 0.2] <- "CTR< 0.2"
ImpSub$CTR.behavior[ImpSub$CTR >= 0.2 &  ImpSub$CTR <0.4] <- "0.2<=CTR <0.4"
ImpSub$CTR.behavior[ImpSub$CTR >= 0.4 &  ImpSub$CTR <0.6] <- "0.4<=CTR <0.6"
ImpSub$CTR.behavior[ImpSub$CTR >= 0.6 &  ImpSub$CTR <0.8] <- "0.6<=CTR <0.8"
ImpSub$CTR.behavior[ImpSub$CTR > 0.8] <- "CTR> 0.4"
``` 

##### **Get the total number of Male, Impressions, Clicks and Signed_In (0=Female, 1=Male)**
```{r count, echo=TRUE, cache=TRUE}

#Gender
sapply(split(ImpSub,ImpSub$Gender),NROW)
#Impressions
sapply(split(ImpSub,ImpSub$Impressions),NROW)
#Clicks
sapply(split(ImpSub,ImpSub$Clicks),NROW)
#Signed_In
sapply(split(ImpSub,ImpSub$Signed_In),NROW)
``` 
<br>

##### **Get the mean of Age, Impressions, Clicks, CTR and percentage of males and signed_In**
```{r mean, echo=TRUE, cache=TRUE}
#Age
mean(ImpSub$Age)
#Impressions
mean(ImpSub$Impressions)
#Clicks
mean(ImpSub$Clicks)
#CTR
mean(ImpSub$CTR)
#%ofmales
mean(ImpSub$Gender/sum(ImpSub$Gender) *100)
#Signed_In
mean(ImpSub$Signed_In)
``` 
<br>

##### **Get the means of Impressions, Clicks, CTR and percentage of males and signed_In  by AgeGroup.**
```{r meanbyagegroup, echo=TRUE, cache=TRUE}
#Impressions
tapply(ImpSub$Impressions,ImpSub$ageGroup,mean)
#Clicks
tapply(ImpSub$Clicks,ImpSub$ageGroup,mean)
#CTR
tapply(ImpSub$CTR,ImpSub$ageGroup,mean)
#%ofmales
tapply((ImpSub$Gender/sum(ImpSub$Gender) *100),ImpSub$ageGroup,mean)
#Signed_In
tapply(ImpSub$Signed_In,ImpSub$ageGroup,mean)
```        
<br>

##### **Create a table of CTRGroup vs AgeGroup counts.**
```{r ctagegroup, echo=TRUE, cache=TRUE}
#CTRGroup vs. AgeGroup
ctragegroup <- table(ImpSub$CTR.behavior,ImpSub$ageGroup)
head(ctragegroup)
```         
<br>

##### **One more plot you think which is important to look at.**
```{r impsubplot4, echo=TRUE, cache=TRUE}
ggplot(subset(ImpSub,CTR>0), aes(x=CTR, colour = ageGroup)) +
  geom_freqpoly(binwidth = 0.015) +
  ggtitle("CTR Click Through Rate by Age Group") + xlab("CTR (Clicks/Impressions)") + ylab("Count")
```         
<br>
